Negative Selection in the 20th Century

### 2020-05-18 why-negative-selection.Rmd
trivia
Add regressions controlling for age at first live birth.

* Use Abdel's 100 PCs, not UKBB's 40. DONE

@@

* Divide siblings regressions by own YOB, into early and late.
  - Use this to confirm the basic "selection decreasing" story
  - Actually almost no change.
  - Overall, diffs between siblings and children regressions
    now look more like ascertainment bias than change over time.

* Plot average scores over time, as an intro graph. DONE
  - Can we work out how to plot selection effects on the "same scale"?
  - Simplest thing would be to use birth dates of children.

@@

* check for new 100 PCs

@@

drake::loadd(famhist)
drake::loadd(rgs)
drake::loadd(score_names)

@@

```{r}
pgs_over_time <- famhist %>% 
      group_by(
        YOB = santoku::chop(YOB, 
          brk_left(seq(1940, 1970, 5), close_end = FALSE), 
          seq(40, 65, 5),
          extend = FALSE
        ) 
      ) %>% 
      filter(! is.na(YOB)) %>% 
      summarize(across(all_of(score_names), mean, na.rm = TRUE)) %>% 
      pivot_longer(-YOB, names_to = "score_name", values_to = "score")

@@

res_pcs %>% 
      mutate(
        estimate = abs(estimate),
        dep.var  = fct_relevel(dep.var, "N children")
      ) %>% 
      ggplot(aes(estimate, dep.var, colour = p.value < 0.05/{{n_regs}})) + 
        geom_jitter(height = 0.15, width = 0) + 
        scale_x_log10(labels = scales::label_number(accuracy = 0.0001)) +
        labs(x = "Effect size (log scale)", y = "") + 
        scale_colour_manual(values = pval_palette)

@@

drake::loadd(res_period)
res_period %<>% 
      tidyr::extract(term, c("score_name", "period"),
        regex="(.*):year_split(.*)")
n_tests <- nrow(res_period)/2

@@

loadd(res_together)
n_regs <- as.double(nrow(res_together))
res_together %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, estimate, order_abs),
        dep.var = recode(dep.var, 
          "n_children" = "N Children", "n_sibs" = "N Siblings")
      ) %>% 
      ggplot(aes(term, estimate, color = p.value < 0.05/{{n_regs}}, shape  = PCs)) + 
        geom_pointrange(aes(ymin = conf.low, ymax = conf.high)) +
        my_hline +
        facet_wrap(vars(dep.var), scales = "free_x") +
        coord_flip() +
        scale_color_manual(values = pval_palette) +
        theme(legend.position = "bottom")
```

@@

>>>>

* Why so many missings for age_fulltime_edu?
  - ah, need f.6138 for those who went to college
  - but note that YearsEdu.ISCED has many fewer missings
  - ask Abdel how it was calculated
  - use this for survey weighting?

@@

* Divide siblings regressions by own YOB, into early and late.
  - Use this to confirm the basic "selection decreasing" story
  - Actually almost no change.
  - Overall, diffs between siblings and children regressions
    now look more like ascertainment bias than change over time.
  - (I checked my code: I hadn't been controlling for one main term when
    running the interaction. Doh.)

* Plot average scores over time, as an intro graph. DONE
  - Can we work out how to plot selection effects on the "same scale"?
  - Simplest thing would be to use birth dates of children. (Maybe average of
    first and last live birth)
  - Also weight?

@@

@@

drake::loadd(rgs)
drake::loadd(score_names)

@@

drake::loadd(famhist)
pgs_over_time <- famhist %>% 
      group_by(
        YOB = santoku::chop(YOB, 
          brk_left(seq(1940, 1970, 5), close_end = FALSE), 
          seq(40, 65, 5),
          extend = FALSE
        ) 
      ) %>% 
      filter(! is.na(YOB)) %>% 
      summarize(across(all_of(score_names), mean, na.rm = TRUE)) %>% 
      pivot_longer(-YOB, names_to = "score_name", values_to = "score")

@@

res_pcs %>% 
      mutate(
        estimate = abs(estimate),
        dep.var  = fct_relevel(dep.var, "N children")
      ) %>% 
      ggplot(aes(estimate, dep.var, colour = p.value < 0.05/{{n_regs}})) + 
        geom_jitter(height = 0.15, width = 0, alpha = 0.9) + 
        scale_x_log10(labels = scales::label_number(accuracy = 0.0001)) +
        labs(x = "Effect size (log scale)", y = "") + 
        scale_colour_manual(values = pval_palette)

@@

drake::loadd(res_period)
res_period %<>% 
      tidyr::extract(term, c("period", "score_name"),
        regex="year_split(.*):(.*)")
n_tests <- nrow(res_period)/2

@@

loadd(res_together)
n_regs <- as.double(nrow(res_together))
res_together %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, estimate, order_abs),
        dep.var = dplyr::recode(dep.var, 
          "n_children" = "N Children", "n_sibs" = "N Siblings")
      ) %>% 
      ggplot(aes(term, estimate, color = p.value < 0.05/{{n_regs}}, shape  = PCs)) + 
        geom_pointrange(aes(ymin = conf.low, ymax = conf.high)) +
        my_hline +
        facet_wrap(vars(dep.var), scales = "free_x") +
        coord_flip() +
        scale_color_manual(values = pval_palette) +
        theme(legend.position = "bottom")
```

@@

## Age at first live birth

```{r corr-age-flb}

drake::loadd(res_age_flb)
drake::loadd(res_sex)

res_sex_comparison <- res_sex %>% 
      filter(sex == "Female") %>% 
      mutate(score_name = term) %>% 
      select(score_name, term, estimate:conf.high)

res_age_flb %<>% filter(term != "age_flb")

res_combined <- bind_rows(
        raw      = res_sex_comparison, 
        with_age = res_age_flb %>% select(score_name, term, estimate:conf.high), 
        .id      = "reg.type"
      ) %>% 
      arrange(score_name)

raw_flb_comparison <- res_combined %>% 
      pivot_wider(
        id_cols     = score_name, 
        names_from  = reg.type, 
        values_from = estimate
      )

cor_raw_flb <- cor(raw_flb_comparison$raw, raw_flb_comparison$with_age)
opp_signed <- sum(
        sign(raw_flb_comparison$raw) != sign(raw_flb_comparison$with_age)
      )

Figure \@ref(fig:plot-age-flb) shows the results of *children* regressions for 
women only,
controlling for age at first live birth. Effect sizes are greatly reduced.
In `r opp_signed` out of `r length(score_names)` cases, they are of the 
opposite sign. The correlation between effect sizes
controlling for age at first live birth, and raw effect sizes, is `r cor_raw_flb`.

```{r plot-age-flb, fig.cap = "Effect sizes on number of children, controlling for age at first live birth (women only). Effect sizes for women without controls are shown for comparison"}

n_regs <- as.double(nrow(res_age_flb))

res_combined %>% 
      mutate(
        score_name = fct_reorder(score_name, estimate, order_abs),
        Regression = ifelse(reg.type == "raw", "Bivariate", "Controlled for age FLB")
      ) %>% 
      ggplot(aes(y = score_name, x = estimate, fill = Regression, 
        color = p.value < 0.05/{{n_regs}})) + 
        geom_vline(xintercept = 0, linetype = "dashed", color = "grey20") +
        geom_point(shape = 21, size = 1.8, stroke = 0.8) + 
        scale_color_manual(values = pval_palette) +
        scale_fill_brewer(type = "qual", palette = 3)
```
